Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho, and Potential Impacts to Local Chinook Salmon Populations

### 2022-03-20 analysis/paper/paper.Rmd
Incorporated edits from reviewers IC and MR

Predation on emigrating juvenile salmonids by piscivorous fishes is a widely-studied source of mortality within reservoirs and dam tailraces in the Columbia River basin. The native Northern Pikeminnow are one of these predators and have been estimated to consume ~8% of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above the reservoir systems. In this study, we examine the abundance of Northern Pikeminnow in a slow-water reach of the Salmon River, Idaho, known as Deadwater Slough, and consider potential impacts of them on local Chinook Salmon populations. The Deadwater Slough contains historically important overwinter rearing habitat and is a migration corridor for Chinook Salmon as well as other Endangered Species Act-listings including steelhead and Sockeye Salmon. We estimated the abundance of Northern Pikeminnow in Deadwater Slough during the peak emigrations of juvenile Chinook Salmon to be 27,874 in the fall and 19,499 in the spring. Using these abundance values, we estimated Northern Pikeminnow consumption of juvenile Chinook Salmon for varying proportions of fish in their diet, ranging from 30% to 90%. Assuming a median scenario where 60% of the Northern Pikeminnow diet is fish, of which 50% is juvenile Chinook Salmon, we estimated the Northern Pikeminnow population can consume 95,120 juvenile Chinook Salmon, annually. Using a sensitivity analysis, the potential impact of predation on juvenile Chinook Salmon under the median scenario is a reduction in adult returns by 539 (95% CI: 231 - 1,335). Given the relatively small size of the Salmon River, the large densities and potential consumption rates of Northern Pikeminnow in Deadwater Slough suggest predation likely has consequential impacts on Chinook Salmon in the Upper Salmon River major population group.

@@

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` km long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough contains historically important overwinter rearing habitat for juvenile Chinook Salmon *O. tshawytscha* and is part of the migratory pathway for upstream Chinook Salmon, the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead. The slough is also inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu* that are potential sources of mortality for these rearing and migrating salmonids.

Quantifying mortality associated with piscine predation in upstream habitats of the Salmon River is of particular interest for spring/summer-run Chinook Salmon (hereafter Chinook Salmon) in the Upper Salmon major population group (MPG). The Salmon River was historically the most productive tributary for Chinook Salmon in the Columbia River basin [@Nemeth1999], but those populations have become depleted in recent decades. Currently, the Upper Salmon MPG supports eight independent, extant populations; this includes Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River populations [@NOAA2017]. The Lemhi River historically contained the largest Chinook Salmon population within the MPG and is therefore a prioritized candidate for restoration of natural processes [@Zimmerman2012].

@@

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey design and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the fall DSR and spring NRR emigrations. During our initial survey in 2019, Northern Pikeminnow were by far the most prevalent piscine predator in Deadwater Slough and consequently became our focal taxa. The initial intent was to survey during the emigration of the 2018 Chinook Salmon brood year (fall 2019 and spring 2020), however logistical constraints during the onset of the Covid-19 pandemic delayed the spring 2020 survey until 2021. Over this period, an additional fall survey occurred during the 2020 DSR emigration. The fall surveys were constrained to two weeks to minimize Northern Pikeminnow immigration/emigration.

Multiple capture methods were employed during the initial fall 2019 survey to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. After evaluating all methods, angling was the most effective method for capturing Northern Pikeminnow while also minimizing potential impacts to ESA-listed adult steelhead that were present during our surveys. Ineffective methods were abandoned in subsequent years and the following analyses will focus on fish captured by angling, unless otherwise noted.

Our study relied heavily on volunteer anglers who were permitted to fish anywhere within Deadwater Slough. Each survey day, anglers boated or hiked their catch (periodically or upon filling a livewell) to a processing station at the boat ramp, approximately 500 m downstream from the top of the slough. For each fish, we recorded the date of capture, species, total length (TL; mm), and whether the fish was previously marked. Unmarked fish were given a physical mark (e.g., hole punch of lower caudal, upper caudal, left pelvic, right pelvic) unique to each day and then were released. Released fish were distributed throughout Deadwater Slough to facilitate mixing back into the population. Fish that died prior to release were included in the total capture (or recapture) count, but not included in the number of marked fish available for recapture after that date. Finally, we recorded the angling start and end time for each crew (person or combinations of persons) to calculate CPUE as the number of Northern Pikeminnow caught per angler hour.

@@

where $M$, $n$, and $m$ are indexed by each survey (day), $i$. With small numbers of recaptures, it may not be appropriate to calculate the standard error of the Schnabel estimator, but the 95% confidence intervals can be calculated using a Poisson approximation [@Krebs1999].<!--The Schnabel estimator is essentially the weighted average of a series of Chapman-modified Lincoln-Peterson estimators.-->

Because captures of new and marked fish were dependent on volunteer anglers, there was concern that sampling was biased to locations preferred by anglers. If marked fish were not recently released or mixed back into those preferred locales, marked fish may have been less susceptible to capture than unmarked fish. To account for this potential bias, we also used an adjusted Schnabel estimator, which assumed that marked and released fish during each survey day were not available for recapture for 48 hours. The adjusted Schnabel estimator, referred to as the "Delayed-mixing" Schnabel estimator, is given by:

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M_{i-1}}{\left(\sum\limits_{i = 1}^k m_i \right) + 1}
$$
All mark-recapture abundance estimators assumed: (1) the population is closed (no immigration, emigration, births or deaths during the survey period), (2) all fish have equal chance of being caught in subsequent survey days, (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, and (5) marks are not missed or mistaken.

To estimate the abundance of Northern Pikeminnow during spring 2021, which was not a mark-recapture design, we used the ratio of mean total CPUE in the fall surveys to the total CPUE in the spring survey, then multiplied that ratio by the mean abundance from fall surveys.

@@

This approach assumes equal capture probabilities for fall and spring surveys. Fall and spring population estimates were converted into linear and areal densities by dividing by the length and area of Deadwater Slough. Length and area were measured in QGIS software [@QGISDevelopmentTeam2022] and using drone-generated orthoimagery.

@@

Anglers reported that Northern Pikeminnow in Deadwater Slough were large-bodied. To describe the population size structure, we calculated the proportional stock density (PSD) for Northern Pikeminnow in Deadwater Slough. PSD is a ratio, expressed as a percentage, between the number of "quality-sized" (larger) individuals and "stock-sized" individuals:

$$
PSD_{i} = 100 * \frac{FQ_{i}}{FS_{i}}
$$

where $FQ_{i}$ is the number of fish $\ge$ quality-length for species $i$, and $FS_{i}$ is the number of fish $\ge$ stock-length for species $i$. For Northern Pikeminnow in Deadwater Slough, we used `r prettyNum(q_size, digits = 0)` mm TL for quality-length and `r prettyNum(s_size, digits = 0)` mm TL for stock-length [@Winther2020].

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes occupying Deadwater Slough. Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks  to be analyzed in a controlled environment. For each stomach sample, wet weight (grams) was recorded for the total stomach contents, including all non-fish items (e.g., macroinvertebrates, organic matter), as well as the portion consisting of fish. Fish and fish remnants were identified to the lowest taxonomic unit using diagnostic bones [@Hansel1988; @Frost2000] or were categorized as unknown. Approximately 5% of Northern Pikeminnow were euthanized for dissection (n = 75) after gastric lavage to validate the efficacy of the methodology.

@@

To estimate the total consumption potential of Northern Pikeminnow in Deadwater Slough during the peaks of fall DSR and spring NRR emigrations, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 applied in R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on the following inputs: predator and prey energy densities, predator start and end weights, and water temperatures. Separate models were run for the fall period (September 15 - November 30) and the spring period (March 1 - May 31) to coincide with peak emigrations of DSR and NRR juveniles from the Lemhi River, the largest Chinook Salmon population in the Upper Salmon MPG.

Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 3,000 J/g for invertebrates and 21,500 J/g for juvenile Chinook Salmon [@Moss2016]. Because we were unable to differentiate juvenile Chinook Salmon from other fish prey, we assume all fish prey have the same energy densities as juvenile Chinook Salmon. The average TL of Northern Pikeminnow caught in Deadwater Slough during our study was converted to fork length (FL) and then to weight (grams) using a weight-length formula from @Parker1995.

@@

The proportion of fish in the Northern Pikeminnow's diet relative to non-fish prey items is unknown in Deadwater Slough. Therefore, we conducted separate model runs that assume different proportions of fish in the diet ranging from 30 - 90%, in 10% increments. These values were supported by observations elsewhere in the Columbia River basin (54-86%: @Shively1996; 48-86%: @Zimmerman1999; 37%: @Gray2001) that suggest fish are the majority of Northern Pikeminnow prey. Fourteen model runs were completed (7 diet scenarios and 2 seasons), each providing an estimate of the total grams of fish consumed by an individual Northern Pikeminnow (Table \@ref(tab:bio-tbl)). To estimate the total biomass of fish consumed by Northern Pikeminnow in Deadwater Slough, we multiplied the grams of fish consumed by an individual Northern Pikeminnow by the estimated Northern Pikeminnow population sizes during fall and spring.

@@

The estimate of total juvenile Chinook Salmon biomass consumed was then converted to the total estimated number of fish consumed each period by dividing by the average weight of DSR (`r avg_DSR_g` g) and NRR (`r avg_NRR_g` g) emigrants in the Upper Salmon MPG. The average weights of DSR and NRR emigrants were calculated from fish captured at seven rotary screw traps upstream of Deadwater Slough during the fall and spring periods. Although the primary impact to Chinook Salmon in Deadwater Slough by Northern Pikeminnow is juvenile predation, the common metric to evaluate salmon recovery is often adult returns. We assessed the potential impacts on adult returns by estimating the number of additional adult Chinook Salmon expected to return to Lower Granite Dam if predation was reduced or eliminated in the Deadwater Slough. This calculation was made by multiplying the total estimated juvenile Chinook Salmon consumed by the median Granite-to-Granite SAR of 0.00614 (sd = 0.00051) from @McCann2019 for Upper Salmon River MPG Chinook Salmon.

@@

Mark-recapture abundance estimates of Northern Pikeminnow in Deadwater Slough during fall 2019 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` and for fall 2020 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` depending on the estimator used (Table \@ref(tab:abund-tab)). For both surveys, the unadjusted multiple-census estimates were larger than the single census estimates and small numbers of recaptures resulted in wide confidence intervals (Figure \@ref(fig:abund-fig)). Because our sampling design most closely matched a multiple census estimator and required a small sample size modification, we deemed the Schnabel estimates most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_F, spring_abund$Uci_F)), big.mark = ","), collapse = " - ")`). Using the delayed-mixing Schnabel estimator, the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`). For spring 2021, we estimated `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm and areal densities (fish/100 m$^2$) of `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively.

>>>>

Predation on emigrating juvenile salmonids by piscivorous fishes is a widely-studied source of mortality within reservoirs and dam tailraces in the Columbia River basin. Native Northern Pikeminnow have been estimated to consume ~8% of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above the reservoir systems. In this study, we examine the abundance of Northern Pikeminnow in a slow-water reach of the Salmon River, Idaho, known as Deadwater Slough, and consider their potential impacts on local Chinook Salmon populations. The Deadwater Slough contains historically important overwinter rearing habitat and is a migration corridor for Chinook Salmon as well as other Endangered Species Act-listings including steelhead and Sockeye Salmon. We estimated the abundance of Northern Pikeminnow in Deadwater Slough to be 27,874 in the fall and 19,499 in the spring, corresponding with the peak emigrations of juvenile Chinook Salmon. Using these abundance values, we estimated Northern Pikeminnow consumption of juvenile Chinook Salmon. Assuming 60% of the Northern Pikeminnow diet is fish, of which 50% is juvenile Chinook Salmon, we estimated the Northern Pikeminnow population can consume 95,120 juvenile Chinook Salmon, annually. Using a sensitivity analysis, Northern Pikeminnow predation on juvenile Chinook Salmon was predicted to reduce adult returns by 539 (95% CI: 231 - 1,335). Given the relatively small size of the Salmon River, the large densities and potential consumption rates of Northern Pikeminnow in Deadwater Slough suggest predation likely has consequential impacts on Chinook Salmon in the Upper Salmon River major population group.

@@

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` km long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough contains historically important overwinter rearing habitat for juvenile Chinook Salmon *O. tshawytscha* and is part of the migratory pathway for upstream Chinook Salmon, the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead. The slough is also inhabited by piscivorous predators, Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*, that are potential sources of mortality for rearing and migrating salmonids.

The Salmon River was historically the most productive tributary for Chinook Salmon in the Columbia River basin [@Nemeth1999], but those populations have become depleted in recent decades. Therefore, quantifying the mortality of spring/summer-run Chinook Salmon (hereafter Chinook Salmon) in the Upper Salmon major population group (MPG) associated with piscine predation is of particular interest. Currently, the Upper Salmon MPG supports eight independent, extant populations; this includes Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River populations [@NOAA2017]. The Lemhi River historically contained the largest Chinook Salmon population within the MPG and is therefore a prioritized candidate for restoration of natural processes [@Zimmerman2012].

@@

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey design and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the fall DSR and spring NRR emigrations. During our initial survey in 2019, Northern Pikeminnow were the most prevalent piscine predator in Deadwater Slough and consequently became our focal taxa. Our intent was to survey during the emigration of the 2018 Chinook Salmon brood year that occurred fall 2019 and spring 2020, however logistical constraints during the onset of the Covid-19 pandemic delayed the spring 2020 survey until 2021. Over this period, an additional fall survey occurred during the 2020 DSR emigration. Fall surveys were constrained to two weeks to minimize Northern Pikeminnow immigration/emigration.

Multiple capture methods were employed during the fall 2019 survey to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. After evaluating all methods, angling was the most effective method for capturing Northern Pikeminnow while also minimizing potential impacts to ESA-listed adult steelhead that were present during our surveys. Less effective methods were abandoned in subsequent years and the following analyses will focus on fish captured by angling, unless otherwise noted.

Our study relied heavily on volunteer anglers who were permitted to fish anywhere within Deadwater Slough. Each survey day, anglers boated or hiked their catch (periodically or upon filling a livewell) to a processing station at the boat ramp, approximately 500 m downstream from the top of the slough. For each fish, we recorded the date of capture, species, total length (TL; mm), and whether the fish was previously marked. Unmarked fish were given a physical mark (e.g., hole punch of lower caudal, upper caudal, left pelvic, right pelvic) unique to each day and then were released. Fish release sites were distributed throughout Deadwater Slough to facilitate mixing back into the population. Fish that died prior to release were included in that day's total capture (or recapture) count, but not included in the number of marked fish available for recapture after that date. Finally, we recorded the angling start and end time for each crew (person or combinations of persons) to calculate CPUE as the number of Northern Pikeminnow caught per angler hour.

@@

where $M$, $n$, and $m$ are indexed by each survey (day), $i$. The 95% confidence intervals were calculated using a Poisson approximation [@Krebs1999].<!--The Schnabel estimator is essentially the weighted average of a series of Chapman-modified Lincoln-Peterson estimators.-->

Because fish capture was dependent on volunteer anglers, there was concern that sampling effort was biased to locations preferred by anglers. If release sites for marked fish were not proximal to preferred angling locales, recently released fish may have been less susceptible to immediate capture than unmarked fish. To account for this potential bias, we included a "Delayed-mixing" Schnabel estimator that assumed fish marked and released on a given survey day were not available for recapture for 48 hours. The "Delayed-mixing" Schnabel estimator, is given by:

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M_{i-1}}{\left(\sum\limits_{i = 1}^k m_i \right) + 1}
$$
All mark-recapture abundance estimators assumed: (1) the population is closed (no immigration, emigration, births or deaths during the survey period), (2) all fish have equal chance of being caught in subsequent survey days, (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, and (5) marks are not missed or misidentified.

Surveys conducted in spring 2021 did not follow a mark-recapture study design so Northern Pikeminnow abundance was estimated as the ratio of mean total CPUE in the fall surveys to the total CPUE in the spring survey, then multiplied that ratio by the mean abundance from fall surveys.

@@

This approach assumes equal capture probabilities for fall and spring surveys. Fall and spring population estimates were converted into linear and areal densities by dividing by the length and area of Deadwater Slough, respectively. Length and area were measured in QGIS software [@QGISDevelopmentTeam2022] and using drone-generated orthoimagery.

@@

Anglers reported that Northern Pikeminnow in Deadwater Slough were large-bodied. To describe the population size structure, we calculated the proportional stock density (PSD) for Northern Pikeminnow in Deadwater Slough. PSD is the percentage of all fish that meet criteria for "quality-length" (larger) individuals and "stock-length" individuals:

$$
PSD = 100 * \frac{FQ}{FS}
$$

where $FQ$ is the number of fish $\ge$ quality-length, and $FS$ is the number of fish $\ge$ stock-length. For Northern Pikeminnow in Deadwater Slough, we used `r prettyNum(q_size, digits = 0)` mm TL for quality-length and `r prettyNum(s_size, digits = 0)` mm TL for stock-length [@Winther2020].

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes occupying Deadwater Slough. Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks  to be analyzed in a controlled environment. For each stomach sample, total wet weight (grams) was recorded for all stomach contents, including all fish and non-fish items (e.g., macroinvertebrates, organic matter). Fish and fish remnants were identified to the lowest taxonomic unit using diagnostic bones [@Hansel1988; @Frost2000] or were categorized as unknown. Approximately 5% of Northern Pikeminnow were euthanized for dissection (n = 75) after gastric lavage to validate the efficacy of the methodology.

@@

To estimate the total consumption potential of Northern Pikeminnow in Deadwater Slough during the peaks of fall DSR and spring NRR emigrations, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 applied in R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on predator and prey energy densities, predator start and end weights, and water temperatures. Separate models were run for the fall period (September 15 - November 30) and the spring period (March 1 - May 31) to coincide with peak emigrations of DSR and NRR juveniles from the Lemhi River, the largest Chinook Salmon population in the Upper Salmon MPG.

Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules (J)/g [@Deslauriers2017]. Prey energy densities were fixed at 3,000 J/g for invertebrates and 21,500 J/g for juvenile Chinook Salmon [@Moss2016]. Because we were unable to differentiate juvenile Chinook Salmon from other fish prey, we assume all fish prey have the same energy densities as juvenile Chinook Salmon. The average TL of Northern Pikeminnow caught in Deadwater Slough during our study was converted to fork length (FL) and then to weight (grams) using a weight-length formula from @Parker1995.

@@

The proportion of fish in the Northern Pikeminnow's diet relative to non-fish prey items is unknown in Deadwater Slough. Therefore, we conducted a series of model runs with varying proportions of fish in the diet ranging from 30 - 90%, in 10% increments. These values were supported by observations elsewhere in the Columbia River basin (54-86%: @Shively1996; 48-86%: @Zimmerman1999; 37%: @Gray2001) that suggest fish are the majority of Northern Pikeminnow prey by volume. A model run was conducted for each combination of diet scenario and season, resulting in fourteen estimates of the total grams of fish consumed by an individual Northern Pikeminnow (Table \@ref(tab:bio-tbl)). To estimate the total biomass of fish consumed by Northern Pikeminnow in Deadwater Slough, we multiplied the grams of fish consumed by an individual Northern Pikeminnow by the estimated Northern Pikeminnow population sizes during fall and spring.

@@

We obtained an estimate of the number of juvenile Chinook Salmon consumed by dividing total biomass consumed by the average weight of DSR (`r avg_DSR_g` g) and NRR (`r avg_NRR_g` g) emigrants in the Upper Salmon MPG. The average weights of DSR and NRR emigrants were calculated from fish captured at seven rotary screw traps upstream of Deadwater Slough during the fall and spring periods. We also quantified the potential impact of Northern Pikeminnow predation on juvenile Chinook Salmon to adult returns by estimating the number of additional adult Chinook Salmon expected to return to Lower Granite Dam if predation by Pikeminnow was reduced or eliminated in the Deadwater Slough. We multiplied the estimated total juvenile Chinook Salmon consumed by the median Granite-to-Granite smolt-to-adult return rate (SAR) of 0.00614 (SD = 0.00051) from @McCann2019 for Upper Salmon River MPG Chinook Salmon.

@@

Mark-recapture abundance estimates of Northern Pikeminnow in Deadwater Slough ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` in fall 2019 and from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` in fall 2020 (Table \@ref(tab:abund-tab)). For both surveys, the unadjusted multiple-census estimates were larger than the single census estimates and small numbers of recaptures resulted in wide confidence intervals (Figure \@ref(fig:abund-fig)). Because our sampling design most closely matched a multiple census estimator and required a small sample size modification, we deemed the Schnabel estimates most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_F, spring_abund$Uci_F)), big.mark = ","), collapse = " - ")`). Using the delayed-mixing Schnabel estimator, the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`). For spring 2021, we estimated `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm and areal densities (fish/100 m$^2$) of `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively.
